---
title: "Problemset 2"
author: "Nikolai German (12712506)"
format: pdf
editor: visual
---
```{r setup}
library(tidyr)
library(dplyr)
library(ggplot2)

```

# Exercise 2.1 MCMC

```{r}
P0 <- matrix(c(1/3, 1/6, 1/6, 1/2, 2/3, 1/2, 1/6, 1/6, 1/3), nrow = 3)
```

## a)

rainy today: $\mathbb{P}(R_{t_0}) = 1$

```{r}
rny_tdy <- c(0, 0, 1)
```

propability of rain on the day after tomorrow, if it's raining today $\mathbb{P}(R_{t_2} | R_{t_0}) = 1$

```{r}
(rny_tdy %*% P0 %*% P0)[[3]]
```

## b)

```{r}
mc <- function(P, state = c("Sunny" = 1, "Cloudy" = 0, "Rainy" = 0), n = 1000) {
  checkmate::assertIntegerish(n, lower = 1, len = 1, any.missing = FALSE)
  checkmate::assertNumeric(state, lower = 0, upper = 1, min.len = 1, any.missing = FALSE)
  checkmate::assertMatrix(P,
                          mode = "numeric",
                          any.missing = FALSE,
                          nrows = length(state),
                          ncols = length(state))
  
  colnames(P0) <- c("Sunny", "Cloudy", "Rainy")
  rownames(P0) <- c("Sunny", "Cloudy", "Rainy")
  
  if (Matrix::rankMatrix(P0)[[1]] < length(state)) {
    stop("P has not full rank!")
  }
  
  E <- eigen(P)
  res <- state %*% E$vectors %*% diag(E$values^n) %*% solve(E$vectors)
  colnames(res) <-c("Sunny", "Cloudy", "Rainy")
  return(res)
}
```


```{r}
E <- eigen(t(P0))
E$vectors[,1] / sum(E$vectors[,1])

mc(P0)
```

## c)

\begin{align}
\pi P &= \pi \\
\iff \pi (P - I) & = 0 \\
\iff (P - I)^{\top} \pi^{\top} &= 0
\end{align}
\begin{equation}
(P - I)^{\top} = 
\begin{bmatrix}
-2/3 & 1/6 & 1/6 \\
1/2 & -1/3 & 1/2 \\
1/6 & 1/6 & -2/3 \\
\end{bmatrix}
\end{equation}

Wir suchen also den Eigevektor zum Eigenwert $\lambda = 1$:

\begin{align}
\begin{bmatrix}
-2/3 & 1/6 & 1/6 \\
1/2 & -1/3 & 1/2 \\
1/6 & 1/6 & -2/3 \\
\end{bmatrix} 
\left( \begin{array}{c}
\pi_1 \\ \pi_2 \\ \pi_3
\end{array} \right)
= 
\left( \begin{array}{c}
0 \\ 0 \\ 0
\end{array} \right) \\
\implies 
\left( \begin{array}{c}
\pi_1 \\ \pi_2 \\ \pi_3
\end{array} \right)
= \alpha * \left( \begin{array}{c}
-1 \\ 1 \\ 0
\end{array} \right)
\end{align}

$$
\pi P = \pi
$$

Wir schreiben den Ausdruck um, zu:
$$
(P - I)^\top \pi^\top = 0
$$


\begin{align}
-2/3 \pi_S + 1/6 \pi_C + 1/6 \pi_R &= 0 \\
1/2 \pi_S -1/3 \pi_C + 1/2 \pi_R &= 0 \\
1/6 \pi_S + 1/6 \pi_C -2/3 \pi_R &= 0
\end{align}

\begin{align}
-4 \pi_S + 1 \pi_C + 1 \pi_R &= 0 \\
3 \pi_S -2 \pi_C + 3 \pi_R &= 0 \\
1 \pi_S + 1 \pi_C -4 \pi_R &= 0
\end{align}


# Exercise 2.2 - MCMC, Gibbs Sampling

## a)

Let $X_i \overset{iid}{\sim} Exp(1)$, $S = \sum_{i=1}^2 X_i$, which leads us to:

\begin{equation}
\mathbb{P}(S \leq s) = \int_0^s f_{X_1}(x) * (F_{X_2}(s - x)) dx
\end{equation}

Taking the complementary propability: $\mathbb{P}(Z > z) = 1 - \mathbb{P}(Z \leq z)$:
\begin{equation}
\mathbb{P}(S > s) = 1 - \int_0^s f_{X_1}(x) * (F_{X_2}(s - x)) dx
\end{equation}

With $f_{X_i} = exp(-x)$ and $F_{X_i} = 1 - exp(-x)$ we obtain:
\begin{align}
\mathbb{P}(S > s) &= 1 - \int_0^s exp(-x) * (1 - exp(-(s - x))) dx \\
&= 1 - \int_0^s exp(-x) + \int_0^s exp(-s) dx \\
&= 1 - F_{X_i}(s) + exp(-s)*s \\
&= 1 - (1 - exp(-s)) + s*exp(-s) \\
&= exp(-s) * (s + 1)
\end{align}

Let's visaualise this probability:

```{r}
Proba_S <- function(x) {
  (x + 1) * exp(-x)
}

plot(Proba_S, xlim = c(0, 11), xlab = "S", ylab = "Probability")
```


And compute $\mathbb{P}(S > 10)$:

```{r}
11 * exp(-10)
```


Compare the frequency of the Sum of two random variables:

```{r}
set.seed(17)
X1 <- rexp(100000)
X2 <- rexp(100000)
S <- X1 + X2

Y <- data.frame(X1, X2, S)[(S > 10), ]

cat(sprintf("Partial of accepted pairs: %.5f", mean(S > 10)))

plot(Y[, c(1,2)])
```

One can easyily see, that the probability for $S > 10$ is very low. Just simulating the random variables would result in a turn-down rate of close to 100%. If the threshold would be even larger, we would discard even more candidates.

To successfully draw from Y, we would need to draw approximatly 2000 times from each X.

## b) 
The Metropolis-Hastings Algorithm constructs a Markov-Chain $Y^t, Y^{t + 1}, ... Y^T$. Given the current state $Y^t$, we draw from a proposal $Q(y|y^t)$. The proposal $y*$ gets then either accepted as $Y^{t+1}$ or rejected.

```{r}
mh <- function(start = c(5, 5), sigma = 1, n = 1000) {
  checkmate::assertNumeric(start, len = 2, any.missing = FALSE)
  checkmate::assertNumeric(sigma, len = 1, any.missing = FALSE, lower = 0)
  checkmate::assertIntegerish(n, len = 1, lower = 1, any.missing = FALSE)
  
  f <- function(y) {
    checkmate::assertNumeric(y, len = 2, any.missing = FALSE)
    if (sum(y) < 10) {
      return(0)
    } else {
      return(dexp(y[[1]]) * dexp(y[[2]]))
    }
  }
  
  alpha <- function(state, proposal) {
    checkmate::assertNumeric(state, len = 2, any.missing = FALSE, lower = 0)
    checkmate::assertNumeric(proposal, len = 2, any.missing = FALSE)
    min(1, f(proposal) / f(state))
  }
  
  Y <- matrix(nrow = n, ncol = 2)
  colnames(Y) <- c("X1", "X2")
  y_t <- start
  i <- 1
  tries <- 0
  
  while (i <= n) {
    proposal <- rnorm(2, mean = y_t, sd = sigma)
    U <- runif(1)
    if (U <= alpha(y_t, proposal)) {
      Y[i, ] <- proposal
      y_t <- proposal
      i <- i + 1
    }
    tries <- tries + 1
  }
  
  structure(Y, acceptance.rate = n/tries)
}
```


```{r}
y <- mh(sigma = 4, n = 200)
y2 <- mh(sigma = .1, n = 200)

plot(y, xlim = c(0, 12), ylim = c(0, 12))
points(y2, col = "red")
```

## c)
Wegen Autokorrelation: akzeptiere nur jeden 40./50. Wert der Markov-Kette (Thinning). Gewährleistet i.i.d.
Startwert beeinflusst die Verteilung enorm (burn-in).
Wahl der Varianz beeinflusst Exploration stark.

## d)


# Exercise 2.3

## a)

The Metropolis-Hasting algorithm states, that $Y^*$ should be accepted, if $U^* < \alpha(Y^*, Y^t)$.
With $U^* \sim Unif(0,1)$ and $\alpha(Y^*, Y^t) = min(1, \frac{f^*(Y^*)}{f*(Y^t)}$, one can clearly see, that the propability of the Markov-Chain to move from state $i$ to state $j$ is $a_{i,j} = min(1, \frac{f^*(Y = j)}{f*(Y = i)}$.

```{r}
f <- function(Y) {
  checkmate::assertCharacter(Y, n.chars = 1)
  checkmate::assertChoice(Y, choices = c("H", "R", "W"))
  switch(Y,
         H = .3,
         R = .1,
         W = .6)
}

alpha <- function(state, proposal) {
  min(1, (f(proposal) / f(state)))
}

states <- c("H", "R", "W")
A <- matrix(nrow = 3, ncol = 3, dimnames = list(states, states))

for (i in seq_len(3)) {
  for (j in seq_len(3)) {
    A[i,j] <- alpha(state = states[[i]], proposal = states[[j]])
  }
}



print(A)
```
## b)

Computing P ist pretty straight forward:
First initializing the Matrix Q for $q = 0.2$:

```{r}
get_Q <- function(q = .2, dim = 3, names = NULL) {
  Q <- matrix(q, nrow = dim, ncol = dim, dimnames = list(names, names))
  diag(Q) <- 1 - (dim - 1) * q
  return(Q)
}

get_Q(names = states)
```

After that, we compute the values $a_{i,j}$ of A, differentiating between the cases $i \ne j$ and $i = j$:

```{r}
Q <- get_Q(names = states)
P <- matrix(nrow = 3, ncol = 3)
for (i in seq_len(3)) {
  for (j in seq_len(3)) {
    if (i != j) {
      P[i,j] <- A[i,j] * Q[i,j]
    } else {
      P[i,i] <- 1 - sum(A[i, seq_len(3) != i] * Q[i, seq_len(3) != i])
    }
  }
}

print(P)
```

## c)

We want to show, that the distribution $\pi = (0.3, 0.1, 0.6)$ is invariant to P, thus beeing an eigenvector of P to the eigenvalue $\lambda = 1$:

```{r}
c(.3, .1, .6) %*% P
```

Since $\pi P = \pi$, it follows directly, that:

\begin{align*}
\pi P^n &= \pi P \cdot P^{n-1} \\
&= \pi P^{n-1} \\
& \vdots \\
&= \pi P \\
&= \pi
\end{align*}

## d)

```{r}
mc2 <- function(start = NULL, states = c("H", "R", "W"), fun = f, q = 0.2, K = 100, burn.in = 0){
  checkmate::assertFunction(fun)
  checkmate::assertCharacter(start, len = 1, any.missing = FALSE, null.ok = TRUE)
  checkmate::assertChoice(start, choices = states, null.ok = TRUE)
  checkmate::assertNumeric(q, len = 1, any.missing = FALSE, lower = 0, upper = .5)
  checkmate::assertIntegerish(K, len = 1, any.missing = FALSE, lower = 1)
  checkmate::assertIntegerish(burn.in, len = 1, any.missing = FALSE, lower = 0, upper = K - 1)
  
  if (is.null(start)) {
    start <- sample(states, 1)
  }
  
  Q <- get_Q(q = q, dim = length(states), names = states)
  
  alpha <- function(state, proposal) {
  min(1, (fun(proposal) / fun(state)))
  }
  
  y_t <- start
  Y <- character(K)
  
  tries <- 0
  i <- 1
  
  while (i <= K) {
    # i.state <- which(states == y_t)
    # probs <- rep(q, length(states))
    # probs[[i.state]] <- 1 - sum(probs[-i.state])
    # proposal <- sample(states, 1, prob = probs)
    proposal <- sample(states, 1, prob = Q[y_t, ])
    U <- runif(1)
    if (U < alpha(y_t, proposal)) {
      y_t <- proposal
      Y[[i]] <- y_t
      i <- i + 1
    }
    tries <- tries + 1
  }
  
  Y <- Y[(burn.in + 1):K]
  structure(Y,
            distribution = table(Y) / (K - burn.in),
            acceptance_rate = K/tries)
}
```


```{r}
samp <- mc2(K = 25000)

tibble(
  H = cumsum(samp == "H"),
  R = cumsum(samp == "R"),
  W = cumsum(samp == "W")
) %>% mutate(n = H + R + W) %>% 
  filter(n > 50) %>%
  pivot_longer(-n, values_to = "count", names_to = "state") %>%
  mutate(freq = count/n) %>%
  ggplot(aes(n, freq, color = state)) +
  geom_point(size = .6, alpha = .3) +
  geom_line() +
  geom_hline(yintercept = .1, lty = "dashed") +
  geom_hline(yintercept = .3, lty = "dashed") +
  geom_hline(yintercept = .6, lty = "dashed") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_light()
  

```





